Using hydraulic geometry and the SWOT satellite to remotely sense river gas exchange velocity

### 2022-01-14 manuscript_v7.Rmd
some ms changes

---
title: "Using hydraulic geometry to remotely sense gas exchange velocity via the SWOT satellite"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Colin J Gleason:
      institute: UMass
  - Christopher J Zappa:
      institute: Lamont
  - Peter A Raymond:
      institute: Yale
  - Merritt H Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts-Amherst
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University
  - Yale: School of the Environment, Yale University
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document:
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx
---

@@

$$FCO_2=k* \delta CO_2 \mathbf{(1)}$$
The structure of equation 1 necessitates that calculations of this flux are highly sensitive to measurements/estimates of *k*. However, *k* can only be directly calculated via a known gas concentration gradient, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020]. In trying to constrain the global fluvial $FCO_2$ flux across millions of rivers, this calculation is impossible, and necessitates the use of predictive models for *k* that are based on easily obtained river hydraulic properties. In that vein, there have been over 20 empirical models developed to predict *k*, generally using some combination of mean velocity $\bar{U}$, shear velocity $U_*$, width $W$, depth $H$, and slope $S$ as predictors [@wangPhysicallyBasedScaling2021]. Note that these models usually predict $k_{600}$, or *k* normalized by a Schmidt number of 600. This is to remove the effect of water temperature and gas type from predictive models, as warmer waters and lower *Sc* numbers both increase gas exchange rates. For example, $k_{600}$ reflects the $k_{CO_{2}}$ at 20 degrees Celsius. Through this normalization, these models exclusively perturb the hydraulic controls on gas exchange [@hallGasExchangeStreams2020].

These empirical models enable estimating a global $FCO_2$ flux ($FCO_2_{upscaled}$) by summing across millions of rivers, but they also change the base parameters that ultimately control that aggregate estimate. That is, by making *k* a function of hydraulics, $FCO_2_{upscaled}$ is now a direct function of river hydraulics. This functional relationship is described in equation 2. It suggests that $FCO_2$ estimates are not only at the mercy of the accuracy and spatiotemporal resolution of $\delta CO_2$, but also the accuracy and resolution of our river hydraulics estimates.

$$FCO_{2_{upscaled}}=f(k_{CO_2}, \delta CO_2)=f(\bar{U}, H, S, W, \delta CO_2) \mathbf{(2)}$$
For example, to upscale $FCO_2$ globally, @raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 all relied on $k_{CO_2}$ values indirectly estimated using mean annual streamflow models and scaling equations to predict hydraulic terms, while @borgesGloballySignificantGreenhousegas2015 used a combination of the above method and a constant $k_{CO_2}$ in space and time to upscale over Africa. In all of these foundational studies, the temporal dynamics of the gas exchange velocity (and thus $FCO_2$) were ignored because of hydraulic data limitations. It has been shown at the field-scale that temporal dynamics of gas exchange can vary widely from site to site [@wallinSpatiotemporalVariabilityGas2011], but it has remained impractical to obtain temporally explicit $k_{CO_2}$ at continental-to-global scales. This knowledge gap is contrasted agaist the robust existing literature exploring spatiotemporal patterns in $CO_2_{water}$ [e.g. @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000].

To that end, the field has made significant progress in estimating $CO_2_{water}$, where we now have river-reach explicit modeling at monthly temporal resolutions using global hydrography datasets [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu etal in review*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking. As equations 1 and 2 dictate, these terms share the burden of calculating $FCO_2$. Therefore, these next-generation models will have to contend with this discrepancy, which is directly dictated by the lack of direct measurements of global river hydraulics.

@@

In theory, this discrepancy between the quality of our $CO_2_{water}$ and $k_{CO_2}$ estimates could be alleviated if direct hydraulics measurements were available at the global scale at a sufficient temporal resolution. This is turn would also address the uncertainty regarding continental-to-global scale temporal dynamics of gas exchange noted earlier. Conveniently, these very data will soon be available via the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. SWOT is expected to launch in 2022 and provide the world's first direct measurements of global water surface extent and elevation at novel temporal resolutions. SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle, measuring rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. Via its direct hydraulic measurements, SWOT is expected to usher in a sea change in global-scale hydrology, and could similarly transform fluvial biogeochemsitry if techniques are developed to ingest SWOT data and infer *k*. Here, we borrow tools from fluvial geomorphology and existing SWOT algorithms to answer the following two questions:

  - Can we leverage the hydraulic geometry of the large rivers SWOT will observe to develop a physically-based model for $k_{600}$ unique to these rivers?

@@

To answer this first question, we use the specific hydraulic geometry of large, SWOT-observable rivers to develop a process-based model for large-river gas exchange. We then take these findings and explore question 2 by implementing the model within an RS of $k_{600}$ algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate'). The goal with BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and should improve results) such that it is completely globally implementable on any river on Earth that SWOT will measure. Finally, we also test whether BIKER can predict actual gas fluxes when paired with an in situ $CO_2$ sensor and compare it against established methods that are reliant on in situ hydraulic measurements.

@@

![Figure 1: Flowchart detailing entire study. Dashed lines indicate workflow to calculate the observed data that we validate against. See Appendix A for variable nomenclature used in this study.](cache/flowchart.jpg)

## 2 Hydraulic geometry in large rivers and predicting fluvial gas exchange velocity
To predict *k* in the very large rivers that SWOT will observe, we start from established process-based models for *k*, impose hydraulic assumptions that are valid in large rivers, and then obtain a suite of models we can test. Following a description of the data used (section 2.1), we outline these established models (section 2.2) and then step through our hydraulic assumptions to arrive at equations that can be ingested by SWOT (section 2.2). Following this model development, all models are empirically tested (section 2.3)

### 2.1 Data
Our gas exchange theory is developed using multiple datasets of directly measured $k_{600}$ and directly measure stream hydraulics collected from the literature. In total, we have collected 763 measurements of $k_{600}$, with 701 of these measurements previously published by @raymondScalingGasTransfer2012 and @ulsethDistinctAirWater2019. These come from tracer studies and thus define $k_{600}$ and the rivers' mean hydraulic properties at the reach scale. The remaining measurements were measured by @churchillPredictionStreamReaeration1964 and @owensReaerationStudiesStreams1964. See Table S1 for a complete list of the studies that collected these measurements.

We expand our dataset of stream hydraulics using a previously published compilation of in situ hydraulic measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019]. That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States, originally taken to calibrate United States Geological Survey (USGS) streamgauge rating curves and made public by the USGS. Ultimately, this dataset was filtered to `r sum(percs$n)` discrete measurements for this study, in addition to the direct measurements in the gas exchange dataset (see Text S1 for our filtering protocol).

### 2.2 Process-based modeling of river *k*
*k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020] and thus extensive field and laboratory experiments have converged on the 'small-eddy model' as first proposed by @lamontEddyCellModel1970 *Zappa said this actually goes back further than this paper* to scale *k* via the smallest-scale turbulent eddies [e.g. @katulStructureFunctionModel2018; @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @wangPhysicallyBasedScaling2021]. This model is provided as equation 3, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number.

$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$
Further, some workers have noted that laboratory and field observations suggest open channel flows with small bed roughness (generally big rivers) do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013]. Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number formulation. *look up other shear reynolds ways to additionally scale k*. This is equation 4 and is referred to as the 'Chainsaw model'. The Chainsaw model is hypothetically useful in low-turbulence flows where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and lead to gas exchange [@talkeTurbulentKineticEnergy2013]. In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-river continuum. Further, to our knowledge the chainsaw model has never been empirically tested in predicting *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a hydraulically-wide river [@talkeTurbulentKineticEnergy2013].

@@

### 2.3 Imposing large-river hydraulic geometry
Rivers and streams change predictably along their longitudinal profile from headwater to ocean, and we can exploit the distinct hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems. In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e. $R_h \approx H$ [@leopoldHydraulicGeometryStream1953]. This is a common assumption in hydraulic and geomorphic modeling of large rivers, and occurs in `r round(percs[percs$flag_swot == 'SWOT' & percs$flag_hydraulicWide == 'Rh=H',]$perc,0)`% of our SWOT-observable river measurements but only `r round(percs_overall[percs_overall$flag_hydraulicWide == 'Rh=H',]$perc,0)`% of non-SWOT observable river measurements (calculated using both datasets of stream hydraulics outlined in section 2.1- see Text S1 for full details). These channels are often referred to as 'hydraulically-wide'.

@@

#### 2.3.2 'Chainsaw' model for gas evasion
We also impose $\epsilon=\epsilon_S$ and $R_H=H$ on the chainsaw model just like we did in section 2.3.1 and arrive at equation 7 (with statistical coefficient $\beta$). The result is identical to the hydraulically-wide small-eddy model (equation 5), except for the exponents which are both steeper. Equation 8 reflects the same derivation, but when $\epsilon=\epsilon_D$ and statistical coefficient $\beta_1$.

@@

|**Description**|**Equation**|**References**|**Empirical test**|**Empirical fit via $r^2$**|
|---------------|------------|--------------|------------------|-----------------|
|Small-eddy, log-law-of-the-wall|$k_{600}=\alpha(gs)^{3/8}H^{1/8}$|@lamontEddyCellModel1970; log-law-of-the-wall|Figure 3a|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)`|
|Small-eddy, form-drag dissipation|$k_{600}=\alpha_1(gS\bar{U})^{1/4}$|@lamontEddyCellModel1970; @moogStreamReaerationNonuniform1999|Figure 3c|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)`|
|Chainsaw, log-law-of-the-wall|$k_{600}=\beta(gs)^{9/16}H^{11/16}$|@moogAirWaterGasTransfer1999; log-law-of-the-wall|Figure 3b|`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eS',]$r2, 2)`|
|Chainsaw, form-drag dissipation|$k_{600}=\beta_1(gs)^{7/16}\bar{U}^{1/4}H^{9/16}$|@moogAirWaterGasTransfer1999; @moogStreamReaerationNonuniform1999|Figure 3d|`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2, 2)`|

### 2.4 Model validation
With equations 5-8 defined, we now empirically test their strength of fit for hydraulically-wide channels. We validate on the dataset of direct measurements of $k_{600}$ and stream hydraulics, after filtering for measurements made in hydraulically-wide channels. This was defined here as those whose hydraulic radius was within 1% of their mean flow depth. All told, this amounts to 166 *double check* direct measurements of hydraulically-wide $k_{600}$ and stream hydraulics to test with. Models were assesed via the coefficient of determination ($r^2$) and results are plotted in Figure 3. Note that Figure 3 axes are plotted in logarithmic space just for visualization: model fit and validation (via $r^2$) were calculated in linear space as their models dictate.

@@

Despite reasonable $r^2$ scores, both small-eddy models produce clearly biased estimates (Figures 3a, 3c). They systematically underestimate lower gas exchange velocities and overestimate higher ones, with the log-law-of-the-wall model (Figure 3a) performing marginally worse both visually and via $r^2$ (`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)` versus `r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)` for Figure 3c). However, when this model is expanded to retard the dissipation rate in the smoothest flows via the Chainsaw model [equations 6 and 7- @moogAirWaterGasTransfer1999], the dynamics are more accurately captured (Figures 3b, 3d) and $r^2$ scores go up appreciably (`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eS',]$r2, 2)` and `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2, 2)` for the log-law-of-the-wall and form-drag dissipation, respectively). Modeling dissipation via total form-drag rather than just via bed shear and the log-law-of-the-wall improves model performance. This is likely because the majority of dissipation occurring in rivers comes not just from the streambed, but from bulk form drag via channel walls, bars, and meanders [@raymondScalingGasTransfer2012; @moogStreamReaerationNonuniform1999; @wangPhysicallyBasedScaling2021; @ulsethDistinctAirWater2019].

The success of the chainsaw model in hydraulically-wide channels, and in particular when coupled with the $\epsilon=\epsilon_D$ model (equation 8, Figure 3d), provides us with a strong physical-model for gas evasion. The river hydraulics terms in equation 8 can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing the gas exchange velocity. This is explored next.

## 3 Exploiting equation 8 to remotely sense gas exchange velocity
We have shown that scaling $k_{600}$ via equation 8 explains `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2,2)*100`% of the variation in $k_{600}$ in hydraulically-wide rivers, which includes nearly all SWOT-observable rivers. Further, equation 8 has only three non-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity. Conveniently, techniques to simultaneously infer hydraulic parameters from SWOT data have been established over the last decade to infer streamflow from SWOT's measurements [e.g. @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018]. For BIKER, we follow the work developed by @durandEstimatingReachaveragedDischarge2014, @hagemannBAMBayesianAMHGManning2017, and @brinkerhoffConstrainingRemoteRiver2020 to infer mean flow velocity and depth from SWOT observations using a modified form of the Manning's equation for mean flow velocity. Following a description of the data used (section 3.1), we detail algorithm development and experimental design (section 3.2) and then we present the validation results (section 3.3).

@@

### 4.3 Towards remotely sensing riverine carbon emissions using SWOT
Figure 7 confirms that BIKER is quite successful, without any in situ information aside from a $CO_2$ logger, at predicting both $FCO_2$ (Figure 6a) and 2) the bulk carbon efflux (Figure 7b-7c). This encouraging result has two main implications for future work coupling remote sensing via SWOT with in situ data. First, it confirms that we can couple BIKER and SWOT with in situ gas concentration loggers to produce reasonable $FCO_2$ estimates at novel temporal resolutions in SWOT-observable rivers. This is particularly useful given recent advances in in situ gas concentration loggers [e.g. @ahoHighestRatesGross2021] but no such similar advances in modeling *k* at equivalent temporal resolutions. BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope (rather than using satellite-based altimeters like SWOT) following recent similar work for estimating streamflow [@harlanDischargeEstimationDense2021].

>>>>

---
title: "Using hydraulic geometry and the SWOT satellite to remotely sense river gas exchange velocity"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Colin J Gleason:
      institute: UMass
  - Christopher J Zappa:
      institute: Lamont
  - Peter A Raymond:
      institute: Yale
  - Merritt H Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts-Amherst
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University
  - Yale: School of the Environment, Yale University
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document:
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx
---

@@

$$FCO_2=k* \delta CO_2 \mathbf{(1)}$$
The structure of equation 1 necessitates that calculations of this flux are highly sensitive to measurements/estimates of *k*. However, *k* can only be directly calculated via a known gas concentration gradient, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020]. In trying to constrain the global fluvial $FCO_2$ flux across millions of rivers, this calculation is impossible, and necessitates the use of predictive models for *k* that are based on easily obtained river hydraulic properties. In that vein, there have been over 20 empirical models developed to predict *k*, generally using some combination of mean velocity $\bar{U}$, shear velocity $U_*$, width $W$, depth $H$, and slope $S$ as predictors [@wangPhysicallyBasedScaling2021]. These models usually predict $k_{600}$, or *k* normalized by a Schmidt number of 600. This is to remove the effect of water temperature and gas type from predictive models, as warmer waters and lower *Sc* numbers both increase gas exchange rates [@hallGasExchangeStreams2020], i.e.e $k_{600}$ reflects the $k_{CO_{2}}$ at 20 degrees Celsius. Through this normalization, these models exclusively perturb the geophysical controls on gas exchange [@hallGasExchangeStreams2020].

These empirical models enable estimating a global $FCO_2$ flux ($FCO_2_{upscaled}$) from millions of rivers, but they also change the base parameters that ultimately control that aggregate estimate. That is, by making *k* a function of hydraulics, $FCO_2_{upscaled}$ is now a direct function of river hydraulics. This functional relationship is described in equation 2. It suggests that $FCO_2$ estimates are not only at the mercy of the accuracy and spatiotemporal resolution of $\delta CO_2$, but also the accuracy and resolution of our river hydraulics estimates.

$$FCO_{2_{upscaled}}=f(k_{CO_2}, \delta CO_2)=f(\bar{U}, H, S, W, \delta CO_2) \mathbf{(2)}$$
For example, to upscale $FCO_2$ globally, @raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 all relied on $k_{CO_2}$ values indirectly estimated using mean annual streamflow models and scaling equations to predict hydraulic terms, while @borgesGloballySignificantGreenhousegas2015 used a combination of the above method and a constant $k_{CO_2}$ in space and time to upscale over Africa. In all of these foundational studies, the temporal dynamics of the gas exchange velocity (and thus $FCO_2$) were ignored because of hydraulic data limitations. It has been shown at the field-scale that temporal dynamics of gas exchange can vary widely from site to site [@wallinSpatiotemporalVariabilityGas2011], but it has remained impractical to obtain temporally explicit $k_{CO_2}$ at continental-to-global scales. This knowledge gap is contrasted agaist the robust existing literature exploring spatiotemporal patterns in $CO_2_{water}$ [e.g. @ahoDistinctConcentrationdischargeDynamics2021; @ahoDifferentialResponseGreenhouse2019; @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000; @liuHydrologicControlsPCO22018a].

To that end, the field has made significant progress in estimating $CO_2_{water}$, where we now have river-reach explicit modeling at monthly temporal resolutions using global hydrography datasets [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu etal in review*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking. As equations 1 and 2 dictate, $CO_2_{water}$ and $k_{CO_2}$ share the burden of calculating $FCO_2$ and therefore these next-generation models will have to contend with this discrepancy, which is directly dictated by the lack of direct measurements of global river hydraulics.

@@

In theory, this discrepancy between the quality of our $CO_2_{water}$ and $k_{CO_2}$ estimates could be alleviated if direct hydraulics measurements were available at the global scale at a sufficient temporal resolution. This is turn would also address the uncertainty regarding continental-to-global scale temporal dynamics of gas exchange noted earlier. Conveniently, these data will soon be available via the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. SWOT is expected to launch in 2022 and provide the world's first direct measurements of global water surface extent and elevation at novel temporal resolutions. SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle, measuring rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. Via its direct hydraulic measurements, SWOT is expected to usher in a sea change in global-scale hydrology, and could similarly transform fluvial biogeochemsitry if techniques are developed to ingest SWOT data and infer *k*. In this context, we borrow tools from fluvial geomorphology and existing SWOT algorithms to answer the following two questions:

  - Can we use hydraulic geometry to develop a physically-based model for $k_{600}$ unique to these rivers?

@@

To answer this first question, we use the specific hydraulic geometry, or the fundamental geomoprhic relationships between streamflow and channel shape [@leopoldHydraulicGeometryStream1953] of large, SWOT-observable rivers to develop a process-based model for large-river gas exchange. We then take these findings and explore question 2 by implementing the model within an RS of $k_{600}$ algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate'). The goal with BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and should improve results) such that it is completely globally implementable on any SWOT-observable river on Earth. Finally, we also test whether BIKER can predict actual gas fluxes when paired with an in situ $CO_2$ sensor and compare it against established methods that are reliant on in situ hydraulic measurements.

@@

![Figure 1: Flowchart detailing the BIKER algorithm. Dashed lines indicate workflow to calculate the observed data that we validate against. See Appendix A for variable nomenclature used in this study.](cache/flowchart.jpg)

## 2 Hydraulic geometry in large rivers and predicting fluvial gas exchange velocity
To predict *k* in the very large rivers that SWOT will observe, we start from established process-based models for *k*, impose hydraulic assumptions that are valid in large rivers, and then obtain a suite of models we can empirically test. Following a description of the data used (section 2.1), we outline these established models (section 2.2) and then step through our hydraulic assumptions to arrive at equations that can be ingested by SWOT (section 2.2). Following this model development, all models are empirically tested (section 2.3)

### 2.1 Data
Our gas exchange theory is developed using multiple datasets of directly measured $k_{600}$ and directly measured stream hydraulics collected from the literature. In total, we have collected 763 measurements of $k_{600}$, with 701 of these measurements previously published by @raymondScalingGasTransfer2012 and @ulsethDistinctAirWater2019. The remaining measurements were measured by @churchillPredictionStreamReaeration1964 and @owensReaerationStudiesStreams1964. See Table S1 for a complete list of the studies that collected these measurements. All measurements come from tracer studies and thus define $k_{600}$ and the rivers' mean hydraulic properties at the reach scale.

To qunatify how many SWOT-obsrvable rivers meet our hydraulic assumptions, we expand our set of stream hydraulics (but not $k_{600}$) using a previously published compilation of in situ hydraulic measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019]. That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States, originally taken to calibrate United States Geological Survey (USGS) streamgauge rating curves and made public by the USGS.

### 2.2 Process-based modeling of river *k*
*k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020] and thus extensive field and laboratory experiments have converged on the 'small-eddy model' as derived by @lamontEddyCellModel1970 and empirically anticipated by @calderbankContinuousPhaseHeat1961. Based on classic surface-renewal theory, this model scales *k* via the smallest-scale (the 'Kolmogorov scale') turbulent eddies and has been repeatedly empirically validated in freshwater systems [e.g. @katulStructureFunctionModel2018; @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @wangPhysicallyBasedScaling2021]. This model is provided as equation 3, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number.

$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$
Further, some workers have noted that laboratory and field observations suggest open channel flows with small bed roughness (generally big rivers) do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013]. Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number formulation. *HERE: Zappa's note that this is frequently done in air-ocean gas exchange studies*. This is equation 4 and is referred to as the 'Reynolds extension' model. The Reynolds model is hypothetically useful in low-turbulence flows where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and lead to gas exchange [@talkeTurbulentKineticEnergy2013]. In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-ocean continuum. Further, to our knowledge this Reynolds extension model has never been empirically tested in predicting river *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a hydraulically-wide river [@talkeTurbulentKineticEnergy2013].

@@

### 2.3 Imposing large-river hydraulic geometry on *k* models
Rivers and streams change predictably along their longitudinal profile from headwater to ocean, and we can exploit the distinct hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems. In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e. $R_h \approx H$. This is a common assumption in hydraulic and geomorphic modeling of large rivers, where the average $\frac{R_h}{H}$ ratio across our `r HG_swot$n` SWOT-observable river flows is `r HG_swot$mean * 100`% (see Text S1 for how we built this dataset). We refer to these flows as 'hydraulically-wide'.

@@

#### 2.3.2 'Reynolds extension' model for gas evasion
We also impose $\epsilon=\epsilon_S$ and $R_H=H$ on the Reynolds extension model just like we did in section 2.3.1 and arrive at equation 7 (with statistical coefficient $\beta$). The result is identical to the hydraulically-wide small-eddy model (equation 5), except for the exponents which are both steeper. Equation 8 reflects the same derivation, but when $\epsilon=\epsilon_D$ and statistical coefficient $\beta_1$.

@@

|**Description**|**Equation**|**Empirical test**|**Empirical fit via $r^2$**|
|---------------|------------|------------------|-----------------|
|Small-eddy, log-law-of-the-wall|$k_{600}=\alpha(gs)^{3/8}H^{1/8}$|Figure 3a|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)`|
|Small-eddy, form-drag dissipation|$k_{600}=\alpha_1(gS\bar{U})^{1/4}$|Figure 3c|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)`|
|Chainsaw, log-law-of-the-wall|$k_{600}=\beta(gs)^{9/16}H^{11/16}$|Figure 3b|`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eS',]$r2, 2)`|
|Chainsaw, form-drag dissipation|$k_{600}=\beta_1(gs)^{7/16}\bar{U}^{1/4}H^{9/16}$||Figure 3d|`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2)`|

### 2.4 Model validation
With equations 5-8 derived, we now empirically test their strength of fit for hydraulically-wide river flows. We validate on the dataset of direct measurements of $k_{600}$ and stream hydraulics, after filtering for measurements made in hydraulically-wide channels (n= `r n`). As equations 5-8 necessitate that this assumption is true, we defined 'hydraulically-wide' as flows whose hydraulic radius was within 1% of their mean flow depth. All told, this amounts to `r n` direct measurements of hydraulically-wide $k_{600}$ and stream hydraulics to test with. Models were assessed via the coefficient of determination ($r^2$) and results are plotted in Figure 3. Note that Figure 3 axes are plotted in logarithmic space just for visualization: model fit and validation (via $r^2$) were calculated in linear space as their models dictate.

@@

Despite reasonable $r^2$ scores, both small-eddy models produce clearly biased estimates (Figures 3a, 3c). They systematically underestimate lower gas exchange velocities and overestimate higher ones, with the log-law-of-the-wall model (Figure 3a) performing marginally worse both visually and via $r^2$ (`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)` versus `r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)` for Figure 3c). However, when this model is expanded to retard the dissipation rate in the smoothest flows via the Reynolds extension model [equations 6 and 7- @moogAirWaterGasTransfer1999], the dynamics are more accurately captured (Figures 3b, 3d) and $r^2$ scores go up appreciably (`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eS',]$r2, 2)` and `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2)` for the log-law-of-the-wall and form-drag dissipation, respectively). Modeling dissipation via total form-drag rather than just via bed shear and the log-law-of-the-wall improves model performance. This is likely because the majority of dissipation occurring in rivers comes not just from the streambed, but from bulk form drag via channel walls, bars, and meanders [@raymondScalingGasTransfer2012; @moogStreamReaerationNonuniform1999; @wangPhysicallyBasedScaling2021; @ulsethDistinctAirWater2019].

The success of the Reynolds extensions model in hydraulically-wide channels, and in particular when coupled with the $\epsilon=\epsilon_D$ model (equation 8, Figure 3d), provides us with a strong physical-model for gas evasion. The river hydraulics terms in equation 8 can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing the gas exchange velocity. This is explored next.

## 3 Exploiting equation 8 to remotely sense gas exchange velocity
We have shown that scaling $k_{600}$ via equation 8 explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of the variation in $k_{600}$ in hydraulically-wide rivers, which includes nearly all SWOT-observable rivers. Further, equation 8 has only three non-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity. Conveniently, techniques to simultaneously infer hydraulic parameters from SWOT data have been established over the last decade to infer streamflow from SWOT's measurements [e.g. @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018]. For BIKER, we follow the work developed by @durandEstimatingReachaveragedDischarge2014, @hagemannBAMBayesianAMHGManning2017, and @brinkerhoffConstrainingRemoteRiver2020 to infer mean flow velocity and depth from SWOT observations using a modified form of the Manning's equation for mean flow velocity. Following a description of the data used (section 3.1), we detail algorithm development and experimental design (section 3.2) and then we present the validation results (section 3.3).

@@

### 4.3 Towards remotely sensing riverine carbon emissions using SWOT
Figure 7 confirms that BIKER is quite successful, without any in situ information aside from a $CO_2$ logger, at predicting both $FCO_2$ (Figure 6a) and 2) the bulk carbon efflux (Figure 7b-7c). This encouraging result has two main implications for future work coupling remote sensing via SWOT with in situ data. First, it confirms that we can couple BIKER with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolutions in SWOT-observable rivers. High temporal fidelity datasets of dissolved $CO_2$ in SWOT-observable rivers are coming online [e.g. @ahoHighestRatesGross2021] but no such similar datasets for $k_{CO_2}$ at equivalent temporal resolutions exist. For rivers unobservable by SWOT, we further suggest that BIKER could be run at the field scale (rather than via satellite-based altimeters like SWOT) using arrays of in situ pressure transducers to estimate water surface slope following recent similar work for estimating streamflow [@harlanDischargeEstimationDense2021]. Regardless, both approaches would produce $CO_2_{water}$ and $k_{CO_2}$ datasets at equivalent temporal resolution and allow us to directly calculate daily to sub-daily $CO_2$ fluxes from river reaches.
